An Overview on Temporal Databases.\\ New Approaches and Proposals.

### 2012-02-24 trunk/book/sources/timeDomain.tex
update for the time domain and the temporal databases

\subsection{Basic concepts and properties}
\label{subsec:basic-concepts}
In information systems, time itself is usually percieved as a linear or cyclic concept~\cite{Jensen94thetsql2}. Therefore, a time domains modelling time is usually represented by a set with an imposed partial order. In general, two main types of time models can be discerned: a \emph{linear} model~\cite{benthem82} and a \emph{cyclic} model~\cite{lorentzos88}. In the linear model, a total order is imposed on the set and the progress of time is seen as a linear matter, while cyclic models are mainly used in the modelling of recurrent processes. It should be noted that the majority of proposals use linear time models.

@@

Thus, an instant is basically an instantanious time point in a time model.

@@

\begin{definition}\textbf{Event}\\
An \emph{\textbf{event}} is an instantaneous fact, i.e. something occurringat an instant.
\end{definition}

@@

In a system, the granularity with the shortest granules is the \emph{chronons granularity}, which is denoted as $\bot$. It is the granularity of which the granules are chronons. It is now possible to define a function $f$, called a \emph{mapping function}, that maps a given granule $i_G$, $i \in \Z$, from a given granularity $G$ to a set of corresponding chronons. The function definition should look like this:
\vspace{-10pt}

\begin{align}
f & : G \rightarrow \Pow(\bot)\nonumber\\
  & : i_{G} \rightarrow \lbrace c_\bot\ |\ (c_\bot\text{ is contained in }i_{G}) \wedge (c_\bot \in \bot)\rbrace\nonumber
\end{align}

For a given granularity $G$ and a given mapping function $f$ from $G$ to $\bot$, it is required that the following properties hold~\cite{Lin97}:

\begin{itemize}
\item
$G$ \emph{is an ordered set}: for a given $i_{G} \in G$ holds: if $c_\bot \in f(i_G)$ and $c^{'}_\bot \in f(j_G)$, then $c_\bot < c^{'}_\bot$ implies that $i_G < j_G$. This is illustrated in figure \ref{fig:granularity-prop1}.
\begin{figure}
\centering
\input{graphs/granularity-prop1.tex}
\caption{$G$ is an ordered set}
\label{fig:granularity-prop1}
\end{figure}

\item
$G$ \emph{is a set of continuous granules}: for a given $i_{G} \in G$ and given chronons $c_\bot$, $c^{'}_\bot$ and $c^{''}_\bot$ holds: if $c_\bot < c^{'}_\bot < c^{''}_\bot$, then $c_\bot \in f(i_G)$ and $c^{''}_\bot \in f(i_G)$ implies that $c^{'}_\bot \in f(i_G)$. This is illustrated in figure \ref{fig:granularity-prop2}.
\begin{figure}
\centering
\input{graphs/granularity-prop2.tex}
\caption{$G$ is a set of continuous granules}
\label{fig:granularity-prop2}
\end{figure}

\item
\emph{The granules in} $G$ \emph{do not overlap}: for given $i_{G}\text{ and } j_{G} \in G$ holds: if $i_G \neq j_G$, then $f(i_G) \cap f(j_G) = \emptyset$. This is illustrated in figure \ref{fig:granularity-prop3}.
\begin{figure}
\centering
\input{graphs/granularity-prop3.tex}
\caption{The granules in $G$ do not overlap}
\label{fig:granularity-prop3}
\end{figure}
\end{itemize}

@@

\begin{definition}
\label{def:coarser-than}\textbf{Coarser Than}
Consider a mapping function $f$ and let $i_G$ and $j_H$ be elements of granularities $G$ and $H$ repectively. Granularity $G$ is now said to be \emph{\textbf{coarser than}} granularity $H$ if:
\begin{equation}
\label{eq:coarser-than}
| f \left( i_G \right)|\ >\ |  f \left( j_H \right)| \nonumber
\end{equation} 
\end{definition}
\vspace{-10pt}
\end{svgraybox}

It is also possible to describe the relation between different granularities. For this, a function, in this chapter denoted as $cast$, may be specified which maps granules from a granularity $G$ to granules from a different granularity $H$. The mathematical notation concerning this function is now:
\begin{equation}
\label{eq:cast-function}
\mbox{cast} \left( i_G,G,H \right) = j_H \nonumber
\end{equation}
This notation indicates that the function $cast$ associates a granule $i_G$ in $G$ to a corresponding granule $j_H$ in $H$. Two kinds of granularity-to-granularity mappings can now be discerned: an \emph{upwards mapping} is a mapping from a granularity $G$ to a coarser granularity $H$, whereas a downwards mapping is a mapping from a granularity $K$ to a finner granularity $L$. Different granularity-to-granularity mappings between several granularities can be represented using a granularity graph, which is a directed graph indicating the mapping conversions. Figure \ref{fig:granularity-graph-example} shows a granularity graph example. Mappings between two granularities may be classified as \emph{regular mappings}, \emph{irregular mappings} or \emph{congruent mappings}.
\begin{itemize}
\item
\emph{Regular mapping}: A regular mapping is a granularity-to-granularity mapping where the mapping function between the two granularities is calculated by means of multiplications and/or divisions and (maybe) an anchor adjustment. E.g. the mapping between hours and minutes is calculated using a multiplication by $60$. Regular mappings are visualised in figure \ref{fig:granularity-graph-example} as normal arrows.
\item
\emph{Irregular mapping}: An irregular mapping is a granularity-to-granularity mapping where the mapping function can not be calculated by means of multiplications and/or divisions. E.g. the mapping between months and days is dependent on the exact month or day. Irregular mappings are visualised in figure \ref{fig:granularity-graph-example} as bold arrows.
\item
\emph{Congruent mapping}: A congruent mapping is a granularity-to-granularity mapping where the two granularities in the mapping have the same granules but a different anchor. E.g. the mapping between the days (Gregorian calendar days) and the academic days. Both concern the same days, but the academic year starts on octobre 1st, whereas the Gregorian calendar year starts on january 1st. Congruent mappings are visualised in figure \ref{fig:granularity-graph-example} as straight lines without arrow heads.
\end{itemize}

@@

%\begin{figure}
%\centering
%\includegraphics[scale=0.5]{graphs/granularityGraph.pdf}
%\caption{A granularity graph example}
%\label{fig:granularity-graph-example}
%\end{figure}

@@

An example of a linear calendar could be the linear hierarchy $\left\{ \text{days, months, years} \right\}$, where days $\sqsubset$ months and months $\sqsubset$ years, together with a validity predicate that excludes time indications like `32 april 2012' or `30 february 2012'.

@@

The Julian Day number \emph{JDN} \cite{Dir96} is a counter. Its value is incremented in one unit every day from 1 January 4713 B.C. at 12:00 noon. The particularity of starting at noon was useful for astronomers: the observations they took one night belonged to the same Julian Day.\\
Note that the Julian Day represents whole days. There is an extension that allows to represent any precision needed (it is called Julian Date). By default, a Julian Day is expressed in Universal Time. (U.T, also known as Solar Time). However, there are representations in Terrestrial Time (T.T), Epheris Time (J.E.D. or J.D.E.). Any time scale different from Universal Time must be explicited after the Julian Day number. There are several conversion formulas~\cite{Usn}\cite{Wik}\cite{Lea} between a date in Gregorian calendar  format and a date in  JDN format. The inverse conversion formula is proposed in~\cite{Fliegel:1968:LEM:364096.364097}\\

>>>>

\subsection{Basic concepts and properties}
\label{subsec:basic-concepts}
In information systems, time itself is usually perceived as a linear or cyclic concept~\cite{Jensen94thetsql2}. Therefore, a time domains modelling time is usually represented by a set with an imposed partial order. In general, two main types of time models can be discerned: a \emph{linear} model~\cite{benthem82} and a \emph{cyclic} model~\cite{lorentzos88}. In the linear model, a total order is imposed on the set and the progress of time is seen as a linear matter, while cyclic models are mainly used in the modelling of recurrent processes. It should be noted that the majority of proposals use linear time models.

@@

Thus, an instant is basically an instantaneous time point in a time model.

@@

\begin{definition}\textbf{Event}\\
An \emph{\textbf{event}} is an instantaneous fact, i.e. something occurring at an instant.
\end{definition}

@@

In a system, the granularity with the shortest granules is the \emph{chronons granularity}, which is denoted as $\bot$. It is the granularity of which the granules are chronons. 
%It is now possible to define a function $f$, called a \emph{mapping function}, that maps a given granule $i_G$, $i \in \Z$, from a given granularity $G$ to a set of corresponding chronons. The function definition should look like this:

\begin{svgraybox}
\begin{definition}
\label{def:mapping-function}
\textbf{Mapping function}~\cite{Lin97}\\ 
A mapping function $f$ is a function that translates from a given granule $i_G$, $i \in \Z$, in a given granularity $G$, to a set of corresponding chronons.\\

\vspace{-10pt}

\begin{align}
f & : G \rightarrow \Pow(\bot)\nonumber\\
  & : i_{G} \rightarrow \lbrace c_\bot\ |\ (c_\bot\text{ is contained in }i_{G}) \wedge (c_\bot \in \bot)\rbrace\nonumber
\end{align}
\end{definition}
\end{svgraybox}

Note that a mapping function $f$ always maps from a given granularity $G$ to the set of chronons $\bot$. Therefore, the output for a mapping function is a subset of $\Pow(\bot)$. \\
 A mapping function $f$ requires that the following properties hold~\cite{Lin97}:

%
%
%For a given granularity $G$ and a given mapping function $f$ from $G$ to $\bot$, it is required that 
\begin{property}
%\item
$G$ \emph{is an ordered set}: for a given $i_{G} \in G$ holds: if $c_\bot \in f(i_G)$ and $c^{'}_\bot \in f(j_G)$, then $c_\bot < c^{'}_\bot$ implies that $i_G < j_G$. This is illustrated in figure \ref{fig:granularity-prop1}.
\begin{figure}
\centering
\input{graphs/granularity-prop1.tex}
\caption{$G$ is an ordered set}
\label{fig:granularity-prop1}
\end{figure}
\end{property}

%\item
\begin{property}
$G$ \emph{is a set of continuous granules}: for a given $i_{G} \in G$ and given chronons $c_\bot$, $c^{'}_\bot$ and $c^{''}_\bot$ holds: if $c_\bot < c^{'}_\bot < c^{''}_\bot$, then $c_\bot \in f(i_G)$ and $c^{''}_\bot \in f(i_G)$ implies that $c^{'}_\bot \in f(i_G)$. This is illustrated in figure \ref{fig:granularity-prop2}.
\begin{figure}
\centering
\input{graphs/granularity-prop2.tex}
\caption{$G$ is a set of continuous granules}
\label{fig:granularity-prop2}
\end{figure}
\end{property}

%\item
\begin{property}
\emph{The granules in} $G$ \emph{do not overlap}: for given $i_{G}\text{ and } j_{G} \in G$ holds: if $i_G \neq j_G$, then $f(i_G) \cap f(j_G) = \emptyset$. This is illustrated in figure \ref{fig:granularity-prop3}.
\begin{figure}
\centering
\input{graphs/granularity-prop3.tex}
\caption{The granules in $G$ do not overlap}
\label{fig:granularity-prop3}
\end{figure}
\end{property}
%\end{itemize}

@@

\begin{definition}
\label{def:coarser-than}\textbf{Coarser Than}
Consider a mapping function $f$ and let $i_G$ and $j_H$ be elements of granularities $G$ and $H$ respectively. Granularity $G$ is now said to be \emph{\textbf{coarser than}} granularity $H$ if:
\begin{equation}
\label{eq:coarser-than}
| f \left( i_G \right)|\ >\ |  f \left( j_H \right)| \nonumber
\end{equation} 
\end{definition}
\vspace{-10pt}
\end{svgraybox}

It is also possible to describe the relation between different granularities. This is called a casting function:
\begin{svgraybox}
\begin{definition}
\textbf{Casting function}~\cite{Lin97}\\
\label{def:casting-fucntion}
A casting function $cast$ maps granules from a granularity $G$ to granules from a different granularity $H$.
\begin{equation}
\label{eq:cast-function}
\mbox{cast} \left( i_G,G,H \right) = j_H \nonumber
\end{equation}
\end{definition}
\end{svgraybox}

This notation indicates that the function $cast$ associates a granule $i_G$ in $G$ to a corresponding granule $j_H$ in $H$. Two kinds of granularity-to-granularity mappings can now be discerned: an \emph{upwards mapping} is a mapping from a granularity $G$ to a coarser granularity $H$, whereas a downwards mapping is a mapping from a granularity $K$ to a finner granularity $L$. Different granularity-to-granularity mappings between several granularities can be represented using a granularity graph, which is a directed graph indicating the mapping conversions.

 Mappings between two granularities may be classified as \emph{regular mappings}, \emph{irregular mappings} or \emph{congruent mappings}.
\begin{itemize}
\item
\emph{Regular mapping}: A regular mapping is a granularity-to-granularity mapping where the mapping function between the two granularities is calculated by means of multiplications and/or divisions and (maybe) an anchor adjustment. E.g. the mapping between hours and minutes is calculated using a multiplication by $60$. 
\item
\emph{Irregular mapping}: An irregular mapping is a granularity-to-granularity mapping where the mapping function can not be calculated by means of multiplications and/or divisions. E.g. the mapping between months and days is dependent on the exact month or day. 
\item
\emph{Congruent mapping}: A congruent mapping is a granularity-to-granularity mapping where the two granularities in the mapping have the same granules but a different anchor. E.g. the mapping between the days (Gregorian calendar days) and the academic days. 
\end{itemize}

\begin{example}
Consider a systems that models both Gregorian Calendar dates as well as Academic dates. The chronons granularity in this system are milliseconds. Figure \ref{fig:granularity-graph-example} shows the complete granularity graph example. The transitions between Chronons and Seconds and between Seconds and Minutes and so on is called a \emph{regular mapping}. The transition between Days and Months is not defined by a multiplication or division, this is called a \emph{irregular mapping}. In the graph, these \emph{irregular mappings} are noted with a bold arrow. Finally, the transition between a Gregorian Calendar Day and an Academic Day is called a \emph{Congruent mapping}: two granularities with the same granules but with different anchors. Both concern the same days, but the academic year starts on October 1st, whereas the Gregorian calendar year starts on January 1st. Congruent mappings are visualised as straight lines without arrow heads.

\end{example}

@@

\begin{figure}
\centering
\includegraphics[scale=0.5]{graphs/granularityGraph.eps}
\caption{A granularity graph example}
\label{fig:granularity-graph-example}
\end{figure}

@@

An example of a linear calendar could be the linear hierarchy $\left\{ \text{days, months, years} \right\}$, where days $\sqsubset$ months and months $\sqsubset$ years, together with a validity predicate that excludes time indications like `32 April 2012' or `30 February 2012'.

@@

The Julian Day number \emph{JDN} \cite{Dir96} is a counter. Its value is incremented in one unit every day from 1 January 4713 B.C. at 12:00 noon. The particularity of starting at noon was useful for astronomers: the observations they took one night belonged to the same Julian Day.\\
Note that the Julian Day represents whole days. There is an extension that allows to represent any precision needed (it is called Julian Date). By default, a Julian Day is expressed in Universal Time. (U.T, also known as Solar Time). However, there are representations in Terrestrial Time (T.T), Epheris Time (J.E.D. or J.D.E.). Any time scale different from Universal Time must be explicit after the Julian Day number. There are several conversion formulas~\cite{Usn}\cite{Wik}\cite{Lea} between a date in Gregorian calendar  format and a date in  JDN format. The inverse conversion formula is proposed in~\cite{Fliegel:1968:LEM:364096.364097}\\
